Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-22 doc/invt.tex
invt.tex: small edits

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \vartheta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \vartheta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  \notag
  %\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{}
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
This serves as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}

@@

We may consider the schedule, $\alpha(t)$,
as a variable transformation of
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited in $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$.
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution is a single exponential decay,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,
then we would have an exponentially decaying schedule.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that completely eliminates certain short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
followed by a perfect filter
for the designated noise modes,
and zero eigenvalues are assigned to these modes.
%
For example,
by assuming
that the PMF is so smooth
that it can be represented by the first $K$ modes,
%
we demand
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0.
  \label{eq:lambda_bandpass}
\end{equation}
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%%
The asymptotic error is then
$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
which is less than that from Eq. \eqref{eq:EA_sbin1}.
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1}
with $t_0$ given by Eq. \eqref{eq:t0_sbin}.
%
%The asymptotic error is given by
%%
%\begin{equation}
%  \Err_A(T)
%  =
%  \frac{   T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{ K - 1 }
%    \Gamma_k
%  ,
%  \notag
%  %\label{eq:error_sinc}
%\end{equation}
%%
%as a generalization of Eq. \eqref{eq:Emin_sbin}
%for the single-bin updating scheme.

@@

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
which recovers the algorithm by
Langfeld, Lucini and Rago\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)/\sigma_z$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined by
Eq. \eqref{eq:maxFk}.
%
This method will be referred to as
the adaptive Gaussian umbrella sampling below.

The use of polynomial basis makes
the bias potential smooth and differentiable,
and hence suitable for MD applications.\cite{zhu2012}
%
However, since we only used two modes
for controlling the mean and variance,
the method only works
for narrow distributions.
%
To sample a wider range,
we can construct an expanded ensemble with
multiple Gaussian distributions
centered at different places via
either the replica exchange
(or parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
%
While the mutual exchange probability
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, \mathcal U(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$\mathcal U(\mathbf x)$ is the unbiased potential energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probability from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{(j)}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(j)}/Z^{(i)}$
from the following approximate formula
%
\begin{align*}
  \ln\frac{ Z^{(j)} }{ Z^{(i)} }
  \approx
  \ln\frac{ \sigma_j }{ \sigma_i }
   -u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.
%
In the $1/t$ scheme,
%
\note{For implementation, we can further simplify
\begin{align*}
  &-u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
   \\
   &=
   (c_1^{(i)} + c_1^{(j)}) \, \Delta z^*
   +
   (c_2^{(i)} - c_2^{(j)}) \, \frac{{\Delta z^*}^2 - 1} {\sqrt 2}
   ,
\end{align*}
where $\Delta z^* = [z_c^{(j)} - z_c^{(i)}]/[\sigma_z^{(j)} + \sigma_z^{(i)}]$.
%
However, the $c_2$ term does not appear to improve histogram flatness. Ideas?
}

@@

As a proof of principle,
we tested the method of adaptive Gaussian umbrella sampling
introduced in Sec. \ref{sec:bandpass}
on the two-dimensional $L\times L$ ten-state
Potts model\cite{wu1982, newman, wang2001, wang2001pre}
along the energy space.
%
We used $L = 32$ and $26$ Gaussian distributions of
the same width $\sigma_E = L$ over $[-1.7L^2, -0.9L^2]$
(the spacing between neighboring Gaussian centers and the width
were roughly the same).
%
For configuration sampling,
although one can use the Metropolis algorithm,
we found it more efficient to use
the Wolff algorithm\cite{wolff1989, newman}
at temperature $c_1/\sigma_E$
followed by a Metropolis acceptance step
to correct for the quadratic potential, $R_2$
(the acceptance ratio was no less than 92\%).
%
The results after $10^5 L^2$ steps are shown in
Fig. \ref{fig:pt_hist}.
%
The Gaussian histograms were evenly spaced with similar width as intended,
and the overall energy histogram was roughly flat,
as shown in \ref{fig:pt_hist}(a).
%
The parameter $c_1/\sigma_E$ roughly represents the local temperature,
and it demonstrated
the backbending behavior\cite{kim2006, *kim2007}
typical in a first-order phase transitions,
as shown in Fig. \ref{fig:pt_hist}(b).
%
The parameter $c_2$ represents the amount of ``pressure''
to restrain the distribution width,
and its value was maximal around the center critical region,
as shown in Fig. \ref{fig:pt_hist}(c).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/pt_hist.pdf}
  }
  \caption{
    \label{fig:pt_hist}
    Application of the adaptive Gaussian umbrella sampling
    on the $32 \times 32$ ten state Potts model.
    %
    (a) Normalized histograms (solid lines) from adaptive Gaussian umbrella sampling
    matched the intended distributions.
    The overall histogram (dashed line) was roughly flat.
    %
    (b) The parameter, $c_1/\sigma_E$, roughly corresponding to
    the statistical temperature, $\beta(E)$,
    demonstrated the backbending behavior.
    %
    (c) The parameter $c_2$ for restraining the variance
    was maximal around the critical region.
  }
\end{center}
\end{figure}

@@

The optimal schedule depends on the updating window function,
which can be limited to a single-bin (as often in the WL case),
or extended to several neighboring bins
taking a Gaussian shape (as often in the metadynamics case).
%
More generally,
each adaptive FDS method is associated with
an updating scheme
that can be mathematically characterized
by an updating matrix.
%
Each column of the matrix gives the normalized
window function for the bin corresponding to the column.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
The eigenvalues represent
the relative rate of error reduction,
and the autocorrelation integrals
represent the intrinsic fluctuation
of the histograms modes.
%
For the single-bin scheme with a set of uniform eigenvalues,
we confirmed that the optimal schedule
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one,
the errors were reduced at different paces in various eigenmodes,
and the optimal schedule,
representing a compromise of the eigenmodes,
is usually not given by a simple closed form,
but is implicitly given by an equation of motion
of a free particle with a position-dependent mass,
Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length
and other parameters.
%
%The initial updating magnitude is always
%half of the previous equilibrium value.
% the value used during equilibration.

In comparing different updating schemes,
we found that
the single-bin updating scheme was optimal
in error reduction
in the long time limit
without a priori assumptions
about the smoothness of the PMF.
%
However, if some eigenmodes
correspond to local noise
that is expected to be absent from the PMF,
we can construct a
class of bandpass updating schemes
that are also optimal for asymptotic convergence.
%
The optimal schedule of these bandpass schemes
is always given by an inverse-time formula.

The Gaussian updating scheme
has the appeal of being able to maintain the smoothness
of the bias potential,
which is important for stable MD applications.
%
However, it requires a careful choice of the width and
the updating magnitude should often be fixed at constant level
for a sufficiently long period before
a magnitude-decreasing schedule could start.
%
For reasonable parameters,
one may use the histogram correction,
Eq. \eqref{eq:uhatav},
in a post-simulation analysis
to regain the possible accuracy loss
and obtain a bias potential similar to
that from the optimized single-bin scheme.
%%
%However, the performance gain
%is relatively small for short simulations
%with a local sampling process such as MD.

>>>>

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \vartheta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \vartheta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{}
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
This serves as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}

@@

We may consider the schedule, $\alpha(t)$,
as a variable transformation of
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited in $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$.
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution is a single exponential decay,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,\footnote{If
  the magntidue of the eigenvalues, $\lambda_k$'s,
  span a wide range, then for a large $\bar q$,
  the sum in Eq. \eqref{eq:mass_func} will be dominated by
  the largest term, where $\lambda_k \approx 1/\bar q$,
  and $M(\bar q) \approx \sqrt{\Gamma_k}/\bar q$.
  }
  %
then we would have an exponentially decaying schedule.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that completely eliminates certain short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
followed by a perfect filter
for the designated noise modes,
and zero eigenvalues are assigned to these modes.
%
For example,
by assuming
that the PMF is so smooth
that it can be represented by the first $K$
modes,\footnote{As an estimate of error,
  we note that truncating a linear or quadratic function
  at mode $K$ (for a non-periodic variable)
  results in an error roughly proportional to
  $|\max_i u_i - \min_i u_i|^2/K^3$.}
we demand
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0.
  \label{eq:lambda_bandpass}
\end{equation}
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%%
The asymptotic error is then
$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
which is less than that from Eq. \eqref{eq:EA_sbin1}.
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1}
with $t_0$ given by Eq. \eqref{eq:t0_sbin}.
%
%The asymptotic error is given by
%%
%\begin{equation}
%  \Err_A(T)
%  =
%  \frac{   T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{ K - 1 }
%    \Gamma_k
%  ,
%  \notag
%  %\label{eq:error_sinc}
%\end{equation}
%%
%as a generalization of Eq. \eqref{eq:Emin_sbin}
%for the single-bin updating scheme.
%

@@

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
which recovers the Langfeld, Lucini and Rago (LLR)
algorithm\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)/\sigma_z$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$.
%
This results in a quadratic bias potential
that can be written as a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This form of bias potential was commonly employed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magnitude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined by
Eq. \eqref{eq:maxFk}.
%
We will refer to this method as
adaptive Gaussian umbrella sampling below.

The use of polynomial basis makes
the bias potential smooth and differentiable,
and hence suitable for MD applications.\cite{zhu2012}
%
However, since we only used two modes
for controlling the mean and variance,
the method can only be applied
to a narrow distribution.
%
To sample a wider range,
we can construct an expanded ensemble with
multiple Gaussian umbrellas
centered at different places via
either the replica exchange
(or parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
%
While the mutual exchange probability
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, \mathcal U(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$\mathcal U(\mathbf x)$ is the unbiased potential energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probability from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{(j)}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(j)}/Z^{(i)}$
from the following approximate formula
%
\begin{align*}
  \ln\frac{ Z^{(j)} }{ Z^{(i)} }
  \approx
  \ln\frac{ \sigma_j }{ \sigma_i }
   -u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.
%
In the $1/t$ scheme,
%
\note{For implementation, we can further simplify
\begin{align*}
  &-u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
   \\
   &=
   (c_1^{(i)} + c_1^{(j)}) \, \Delta z^*
   +
   (c_2^{(i)} - c_2^{(j)}) \, \frac{{\Delta z^*}^2 - 1} {\sqrt 2}
   ,
\end{align*}
where $\Delta z^* = [z_c^{(j)} - z_c^{(i)}]/[\sigma_z^{(j)} + \sigma_z^{(i)}]$.
%
However, the $c_2$ term does not appear to improve histogram flatness. Ideas?
}

@@

As a proof of principle,
we tested the adaptive Gaussian umbrella sampling method
introduced in Sec. \ref{sec:bandpass}
on the two-dimensional $L\times L$ ten-state
Potts model\cite{wu1982, newman, wang2001, wang2001pre}
in the energy space.
%
We spread the Gaussian umbrellas
the same width $\sigma_E = L$ over $[-1.7L^2, -0.9L^2]$
with a spacing of $2L$.
%
For configuration sampling for each umbrella,
we used the Wolff algorithm\cite{wolff1989, newman}
(which appeared to be more efficient than the Metropolis algorithm),
at temperature $c_1/\sigma_E$
followed by a Metropolis step with
acceptance probability
$\min\bigl\{1, e^{c_2 [R_2(E) - R_2(E')]} \bigr\}$,
whose average was no less than 92\%.
%
To ensure stability,
we avoid transitions among umbrellas
until the current umbrella has
completed the WL stage-wise way of
reducing the updating magnitude
and started the inverse-time schedule.\cite{belardinelli2007,
*belardinelli2007jcp, *belardinelli2008, *belardinelli2016}

The results for the $L = 64$ case
after $10^5 L^2$ steps are shown
in Fig. \ref{fig:pt_hist}.
%
The Gaussian histograms were evenly spaced with similar width as intended,
and the overall energy histogram was roughly flat,
as shown in Fig. \ref{fig:pt_hist}(a).
%
The parameter $c_1/\sigma_E$ roughly represents the local temperature,
and it demonstrated
the backbending behavior\cite{kim2006, *kim2007}
typical in a first-order phase transitions,
as shown in Fig. \ref{fig:pt_hist}(b).
%
The parameter $c_2$ represents the amount of ``pressure''
to restrain the distribution width,
and its value was maximal around the center critical region,
as shown in Fig. \ref{fig:pt_hist}(c).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/pt_hist.pdf}
  }
  \caption{
    \label{fig:pt_hist}
    Application of adaptive Gaussian umbrella sampling
    on the $64 \times 64$ ten state Potts model.
    %
    (a) Normalized histograms (solid lines) from adaptive Gaussian umbrella sampling
    matched the intended distributions.
    The overall histogram (dashed line) was roughly flat.
    %
    (b) The parameter, $c_1/\sigma_E$, roughly corresponding to
    the statistical temperature, $\beta(E)$,
    demonstrated the backbending behavior.
    %
    (c) The parameter $c_2$ for restraining the variance
    was maximal around the critical region.
  }
\end{center}
\end{figure}

@@

The optimal schedule depends on the updating window function,
which can be limited to a single-bin (as often in the WL case),
or extended to several neighboring bins
taking a Gaussian shape (as often in the metadynamics case).
%
We can generally characterize
an updating scheme with an updating matrix,
with each column being the normalized
window function for the bin of the column.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
which represent the relative rates of error reduction,
and the integrals of the autocorrelation functions,
which represent the intrinsic histogram fluctuation,
of the eigenmodes.
%
For the single-bin scheme with a set of uniform eigenvalues,
we confirmed that the optimal schedule
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one,
the optimal schedule, having to balance
the different rates of error reduction of the eigenmodes,
is usually not given by a simple closed form,
but is implicitly given by an equation of motion
of a free particle with a position-dependent mass,
Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length
and other parameters.

In particular, the Gaussian updating scheme is often
equipped with a set of rapidly diminishing eigenvalues.
%
While this feature helps suppressing
the accumulation of random error and
maintaining the smoothness of the bias potential,
it also slows down the reduction
of the initial bias, if any,
in the middle-range eigenmodes.
%
Thus, one needs to carefully choose the width of the Gaussian
and the optimal schedule of decreasing the updating magnitude,
which sometimes needs to be fixed at a maximal acceptable level
for a long period before any reduction can happen.
%
The possible efficiency loss in
error reduction of the long- and middle-range modes
can often be compensated
by the histogram-based correction formula, Eq. \eqref{eq:uhatav},
for post-simulation analysis,
which ideally would remove the remaining bias
and recover a bias potential with precision
similar to that from the single-bin case.

In comparing different updating schemes,
we found that
the single-bin updating scheme was optimal
in error reduction
in the long time limit
without a priori assumptions
on the smoothness of the PMF.
%
This scheme can be generalized to
a class of bandpass updating schemes,
which are also asymptotically optimal,
but completely filter the components
in the noise modes.
%
The optimal schedule of the bandpass schemes
is always given by an inverse-time formula.
%
If the bandpass updating scheme was constructed
from orthogonal polynomials,
the updating scheme became a generalization
of the LLR algorithm\cite{langfeld2012, *pellegrini2014},
and it could be used for parameter adjustment
in umbrella sampling with quadratic
bias potentials\cite{neuhaus2006, *neuhaus2007, zhu2012}.
